Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-03 submission/manuscript.Rmd
begin editing chipseq section text

### KDM5C-mediated removal of H3K4me3 during early embryogenesis promotes long-term repression of germline genes

**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
	A) KDM5C binding in EpiLCs vs PNCs - number of peaks and percentage of bound genomic regions
	B) KDM5C binding in Epilcs vs pncs germline genes - GO 
	C) Left - bar graph of germline DEGs KDM5C is bound to in EpiLCs Right 
	D) Example bigwigs
		* Brain and EpiLC shared, EpiLC-specific, Brain-specific, 
	E) motif analysis of KDM5C-bound vs unbound genes

	--- separate into two figures
	D) Bigwigs of Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
	E) Increase in H3K4me2 in EpiLCs at germline genes (left - bigwigs, right average signal at germline gene TSS? Just KDM5C-bound genes?)
	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later). 
	F) KDM5C expression ESC --> EpiLC (increasing then decreasing) Left - RNA right - protein 
	G) RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
	H) DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)

	Direct vs indirect binding motif analysis

Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription. However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including key regulators of meiosis and germline identity. While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other germline drivers that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.

@@

@@

To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`kb from TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A). Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B). Contrastingly, germline ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).

@@

----

>>>>

### KDM5C binds to a subset of germline gene promoters during early embryogenesis

- **note: do Direct vs indirect DEGs motif analysis**
- However, it is currently unclear if KDM5C binds to all germline DEGs and if its binding is maintained at any germline genes in neurons.

Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress hippocampal germline DEGs and KDM5C is not bound to their promoters in neurons[@scandagliaLossKdm5cCauses2017]. There is some evidence KDM5C binds to select germline gene promoters in ESCs[@scandagliaLossKdm5cCauses2017], including two recent independent screens that found KDM5C binds to Dazl's promoter[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. As KDM5C's binding at germline gene promoters has not been systematically characterized, it is currently unclear what types of germline genes KDM5C regulates and if its binding is maintained at any germline genes in neurons.

@@

GO_KDM5C_ChIP <- read.csv("../results/ChIPseq_KDM5C_GO_EpiLC_vs_PNC_unique.csv")

@@

To further characterize the mechanism behind _Kdm5c_-KO germline gene misexpression, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016]. EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`kb from TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).

- PNC EpiLC shared unique be come explicit (Figure 4B)

While the majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs, many genes were not

a subset of promoter-bound genes were shared between EpiLCs and PNCs

- We then compared the known functions of genes with KDM5C bound to their promoter in EpiLCs vs PNCs.

When we compared the functions of genes with KDM5C bound to their promoters, germline-specific ontologies were only enriched in EpiLCs (Figure 4B).

Ggermline-specific ontologies were only enriched in EpiLCs (Figure 4B).   
When we compared the functions of genes with KDM5C bound to their promoters,

Ontologies enriched EpiLCs included including meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)` and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3).

Contrastingly, germline-specific ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).

@@

### KDM5C erases H3K4me3 to promote long-term repression of germline genes via DNA methylation
